The role of aneuploidy in the evolution of cancer drug resistance

### 2024-01-23 EvolutionaryRescue1.tex
add comments to manuscript

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to various treatments. 
One evolutionary mechanism by which a population of cancer cells can to adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we examine how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. We also approximate the mean recurrence time for the tumor to revert back to its initial size.
We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
% TODO missing something here...
\end{abstract}

@@

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
\subsection*{Evolutionary model}

@@

We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division. The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division. Given that only about one to two chromosome additions are ``beneficial'' for cancer resistance (see details next paragraph) we choose $u=10^{-2}$ per cell division. For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ (i.e.  $\tilde{u}=10^{-3}$ per cell division) as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases (\Cref{rescue_threshold}B). Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A). Increasing the mutation rate $v$ decreases the effect of aneuploidy and increases the probability of evolutionary rescue.

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We note that for this parameter regime we have $N_a^*\approx4.4\times10^6$ and $N_m^*\approx4\times10^7$ and this allows us to define the following categories: small tumors have size $N<N_a^*$, intermediate sized tumors have $N_a^*<N<N_m^*$ and large tumors have size $N>N_m^*$ . For the remaining of this paper we use these definitions when referring to small, intermediate or large tumors. Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Density-dependent growth.}

@@

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes. Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \left(v\lambda_wNp_m\right)^{-1}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}). Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time. This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times. Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e. $\tau_a\approx1/\abs{\Delta_a}$. For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$). Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation. Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.}

In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant. We note that $\tau_a^r$ has the following asymptotic behavior:
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}). Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time. In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size. Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse. Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}

In Appendix E we derive the probability that by time $t$ a successful mutant has been generated. \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.  We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.

@@

We hypothesized that presence of \emph{standing variation} 
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. However, we observe that for reasonable parameter values evolutionary rescue is more likely to occur through de-novo aneuploidy (\Cref{rescue_denovo}). If the the fraction of the tumor cells which have the ``beneficial'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$ then evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells. In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}).

\paragraph{Experimental future direction}
Our model's predictions may be tested by experiments \citep{martin2013probability}. For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present. We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Josh's comments

% To me the big picture of the paper isn't clear.  I appreciate that the model can replicate what one expects, but what are the clinical implications, or surprising results?  Which of the scenarios are most realistic given the available clinical and experimental data?  What are the biological interpretations of the derived formulate

% it's also not clear to me from the introduction that this is a novel model, which it is, no?

% Big picture:
% (1) Any time you say "we observe" or "interestingly" it sets up the expectation for biological interpretation - going through the paper and ensuring that such interpretation is present will help a lot
% (2) What's the overall story of the paper beyond we made a model and it does what we expect.  I'm not saying that there isn't more to the story here, I think there is, but it's not immediately clear to me what that more is.
% (3) adding interpretation to figure discussion/caption.  For example, in figure 3  (to me, but perhaps I'm wrong) it is unsurprising that the probability of rescue increases with population size, and that partially resistant saturates faster than stationary which is faster than partially tolerant.  It seems to me the main message of the figure right now is that the presence of aneuploidy in turmor cells, even if only partially tolerant to treatment, facilitates rescue. However, it seems like the more crucial message in the figure is the number of cells at which the curves saturates.  Perhaps one could plot the saturation level in terms of number of cancer cells as a function of the parameter value(s) which define each scenario. or perhaps some combination of the N*s and the saturation points

% other comments:
% figures 2, 3 shouldn't have the same bolded caption 
%

% some line item comments are below
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to various treatments. 
One evolutionary mechanism by which a population of cancer cells can to adapt to chemotherapy is aneuploidy.  
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we examine how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. We also approximate the mean recurrence time for the tumor to revert back to its initial size.
We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
% TODO missing something here...
\end{abstract}

@@

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
% It feels like conclusions are missing from this paragraph, we did these things and found what?...
% from the abstract "aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size." is this not already suggested by the sources in paragraph two above?
% This paragraph should make clear the contribution of this paper, to me at the moment it's not...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
\subsection*{Evolutionary model}

@@

We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division. % the above sentence feels repetitive 
The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division. Given that only about one to two chromosome additions are ``beneficial'' for cancer resistance (see details next paragraph) we choose $u=10^{-2}$ per cell division. For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ (i.e.  $\tilde{u}=10^{-3}$ per cell division) as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
% the bracket here is meant to denote multiplication, approximation, what?
% Can you interpret this more directly in a biological sense?
In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases (\Cref{rescue_threshold}B). Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A). Increasing the mutation rate $v$ decreases the effect of aneuploidy and increases the probability of evolutionary rescue.

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We note that for this parameter regime we have $N_a^*\approx4.4\times10^6$ and $N_m^*\approx4\times10^7$ and this allows us to define the following categories: small tumors have size $N<N_a^*$, intermediate sized tumors have $N_a^*<N<N_m^*$ and large tumors have size $N>N_m^*$ . For the remaining of this paper we use these definitions when referring to small, intermediate or large tumors. Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}. % might just be because I'm not familiar with this area of biology, but can this be interpreted further? i.e. what are the consequences, if any, of this?
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Density-dependent growth.}

@@

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes. Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \left(v\lambda_wNp_m\right)^{-1}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}). % might just be because I'm not familiar with this area of biology, but why is this interesting, does it defy a priori expectations?
Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time. This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times. Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e. $\tau_a\approx1/\abs{\Delta_a}$. For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$). Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation. Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.} % and? whats the further interpretation of this?

In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant. We note that $\tau_a^r$ has the following asymptotic behavior:
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}). Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time. In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size. Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse. Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment. % this is good, more things like this!
\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}

In Appendix E we derive the probability that by time $t$ a successful mutant has been generated. \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.  We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days. % and? so what...

@@

We hypothesized that presence of \emph{standing variation} 
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. However, we observe that for reasonable parameter values evolutionary rescue is more likely to occur through de-novo aneuploidy (\Cref{rescue_denovo}). If the the fraction of the tumor cells which have the ``beneficial'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$ then evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells. In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}). % maybe pull some of this out for the last paragraph of the introduction?

\paragraph{Experimental future direction}
Our model's predictions may be tested by experiments \citep{martin2013probability}. For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present. We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.   
% I may need to reread, but what are the specific predictions in this paper that can be tested by experiments?
